Non-crop sources of beneficial arthropods vary within-season across a prairie agroecosystem

### 2020-11-10 manuscript/manuscript.Rmd
Updated manuscript
- Moved abstract to separate file
- Moved Supplemental to separate file

---
title: "Non-crop sources of beneficial arthropods vary within-season across a prairie agroecosystem"
author:
  - name: Samuel Robinson
    email: samuel.robinson@ucalgary.ca
    affiliation: University of Calgary
    footnote: 1
  - name: Diane Edwards
    email: diane@abienvironmental.ca
    affiliation: Ambrose University
  - name: Jess Vickruck
    email: jessica.vickruck@canada.ca
    affiliation: University of Calgary
  - name: Lincoln Best
    email: lincoln.best@oregonstate.edu
    affiliation: Oregon State University
  - name: Paul Galpern
    email: paul.galpern@ucalgary.ca
    affiliation: University of Calgary
address:
  - code: University of Calgary
    address: 2500 University Drive NW, Calgary, AB
  - code: Ambrose University
    address: 150 Ambrose Circle SW, Calgary, AB
  - code: Oregon State University
    address: 1500 SW Jefferson Way, Corvallis, OR
footnote:
  - code: 1
    text: "Corresponding Author"
date: "Spring 2020"
abstract: "Ground-dwelling arthropods can be important generalist predators in agroecosystems, and can use non-crop features as overwintering habitats. However, it is unclear which types of landscape features constitute useful non-crop habitat, and at what spatial scale organisms gather resources. Additionally, the same landscape feature may act as a source or destination for arthropods at different times of the year, but this is rarely considered. We modeled the abundance of four highly-abundant species of Canadian prairie arthropods caught in pitfall traps, and used functional regression to simultaneously consider both their habitat preferences and the timing of sources and destinations. _Pterostichus melanarius_ (Coleoptera: Carabidae) and _Pardosa moesta_ (Arachnida: Lycosidae) appear to be attracted to canola (_Brassica napus_ L.) during the early summer, then dispersed to grasslands, wetlands, and grassy road margins during the end of the summer. This suggests that grasslands and wetlands act as seasonal reservoir habitats for predatory arthropods, and that farmers should consider preserving existing habitat in order to maintain pest-control services across the season."

bibliography: refs.bib
csl: ecology-letters.csl
geometry: "left=3cm,right=3cm,top=3cm,bottom=3cm"
output: 
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    toc: false
    fig_caption: true
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0} \newpage \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{setspace}
  \linenumbers
---

<!-- This argument in YAML header adds "manuscript_appendix.md" after document -->
<!-- output: -->
<!--   bookdown::html_document2: -->
<!--     pandoc_args: ["--include-after-body=manuscript_appendix.md"] -->

@@

Key words: beetles; spiders; harvestmen; movement; natural enemies; functional regression

@@

## Data collection
We used a set of `r sum(NtrapsPer$n)` pitfall traps installed in road margins (minimum of 5 m away from the road edge, `r with(NtrapsPer, sum(n[trapLoc=='ditch']))` traps) and in-field locations (`r with(NtrapsPer, sum(n[trapLoc!='ditch']))`) across southern Alberta, Canada in 2017 (Figure \@ref(fig:siteMap)).
The sites spanned a west-to-east gradient of four natural subregions, including foothills parkland, foothills fescue, mixedgrass, and dry mixedgrass [@ABRegions2006].
Traps were placed starting on `r format(as.Date(as.character(min(PteMelMod$tempTrap$startjulian)),format='%j'),format='%B %d')`, and collections ended on `r format(as.Date(as.character(max(PteMelMod$tempTrap$endjulian)),format='%j'),format='%B %d')`, with collection occurring continuously and traps being emptied every `r round(mean(PteMelMod$tempTrap$trapdays),1)` days on average (SD: `r round(sd(PteMelMod$tempTrap$trapdays),1)`). 
This resulted in `r nrow(PteMelMod$tempTrap)` unique collection events, taking place across a total of `r round(sum(PteMelMod$tempTrap$trapdays),1)` trapping days.
In-field traps were placed in canola crops (`r with(NtrapsPer, sum(n[trapLoc=='canola']))`), wetlands (`r with(NtrapsPer, sum(n[trapLoc=='wetland']))`), grassy field edges (`r with(NtrapsPer, sum(n[trapLoc=='pivot']))`), and remnant prairie grasslands (`r with(NtrapsPer, sum(n[trapLoc=='native']))`).
Traps in canola were installed at 25, 75, and 200 m along a transect heading away from the nearest non-crop feature (wetland, grassy field edges, or remnant prairie), while the trap at 0 m was installed in the non-crop feature itself.
We used 582 mL Solo® cups buried up to the rim and partially filled with propylene glycol, with 2 cm wire mesh mounted over the rim to prevent vertebrates from falling into the traps.
Specimens were identified to species using appropriate taxonomic literature [@dondale1990; @edgar1990; @vogel2004; @yigit2007].

@@

Functional regression [@ramsay2004; @yen2014] was used to incorporate landscape information at different distances, allowing assessment of both local and regional landscape composition [@galpern2020].
Scalar-on-function regression is a special type of linear regression model ($y=X\beta + e$), where the columns of the model matrix $X$ contain some continuous predictor of the scalar $y$, and the values of the coefficients ($\beta$) are modeled as a smooth function $f(x)$ of the predictors.
<!-- As an example of scalar-on-function regression, biomass in a plot (scalar) could be modeled as a smoothed function of monthly average temperatures at that plot (function), meaning that the effect of monthly temperatures would also be a function of time. -->
In our case, the proportion of landscape cover ($X$) within each annulus surrounding the trap is a function of distance away from the trap, meaning that the coefficients are a smoothed function of distance ($\beta_i = f(distance_i)$), and represent the additive effect of a given type of landscape cover at distances away from the trap (i.e. $X\beta$).
This allows for the possibility that the size of the landscape "grain" relevant to a given organism may change over the course of its life stages [@addicott1987; @lima1996; @gardiner2010].
It also allows for the possibility that certain land cover types may be locally beneficial, but detremental at large scales, indicating that the cover type may not constitute a completely usable habitat.

@@

# Applications

@@

@@

Many thanks to the farmers who gave us access to their land for this study. This also would not have been possible without the help of Michael Gavin, Jennifer Retzlaff, and Mugunzal Abarmayan, and John Swann. Funding for this research was provided by Ducks Unlimited Canada's Institute for Wetland and waterfowl Research, the Alberta Canola Producers Commission, Manitoba Canola Growers Association, SaskCanola, the Alberta Biodiversity Monitoring Institute, the Alberta Conservation Association.

\newpage

# Supplementary Material {-}

\beginsupplement

```{r concurvity, echo=FALSE, fig.cap='Concurvity estimates for reduced cover classes used in models.', out.width='100%'}  
  include_graphics('../figures/coverCorPlots/concurvityEstimate_reduced.png',dpi=NA)
```

```{r landscapeComp, echo=FALSE, fig.cap='Percent cover of landscape cover classes in annuli surrounding each trap location. Sites are represented by individual lines. Mean cover for each class is listed in each sub-heading.', fig.width=5, fig.height=5}  
coverList %>% group_by(cover) %>% mutate(p=sum(prop)) %>% ungroup() %>% mutate(p=100*round(p/sum(unique(p)),4)) %>% 
  mutate(cover=paste0(as.character(cover),' (',p,'%)'),prop=prop*100) %>% 
  mutate(cover=factor(cover,levels=unique(cover))) %>% 
  ggplot(aes(dist,prop,group=site))+geom_line(alpha=0.2)+
  facet_wrap(~cover,ncol=4)+theme_classic()+theme(axis.text = element_text(size = 8),strip.text=element_text(size=8))+
  labs(y='Percent cover',x='Distance from trap location (m)')
```

(ref:PteMelSTCap) Temporal and spatial components of _Pterostichus melanarius_ activity-density (after accounting for landscape composition and trap location). a) shows the effect of day of year, and b) shows the spatially smoothed effect of site location.

```{r PteMelSpatiotemporal, echo=FALSE, fig.cap='(ref:PteMelSTCap)', out.width='100%'}
  include_graphics('../figures/Pterostichus_melanarius_raneff.png',dpi=NA)
```

(ref:ParDisSTCap) Temporal and spatial components of _Pardosa distincta_ activity-density (after accounting for landscape composition and trap location). a) shows the effect of day of year, and b) shows the spatially smoothed effect of site location.

```{r ParDisSpatiotemporal, echo=FALSE, fig.cap='(ref:ParDisSTCap)', out.width='100%'}
  include_graphics('../figures/Pardosa_distincta_raneff.png',dpi=NA)
```

(ref:ParMoeSTCap) Temporal and spatial components of _Pardosa moesta_ activity-density (after accounting for landscape composition and trap location). a) shows the effect of day of year, and b) shows the spatially smoothed effect of site location.

```{r ParMoeSpatiotemporal, echo=FALSE, fig.cap='(ref:ParMoeSTCap)', out.width='100%'}
  include_graphics('../figures/Pardosa_moesta_raneff.png',dpi=NA)
```

(ref:OpilioSTCap) Temporal and spatial components of _Phalangium opilio_ activity-density (after accounting for landscape composition and trap location). a) shows the effect of day of year, and b) shows the spatially smoothed effect of site location.

```{r OpilioSpatiotemporal, echo=FALSE, fig.cap='(ref:OpilioSTCap)', out.width='100%'}
  include_graphics('../figures/Opiliones_raneff.png',dpi=NA)
```

```{r PteMelLinear, echo=FALSE, warning=FALSE}
pmelLinTerms %>% select(-p) %>% 
  kable(.,caption='Trap location (cover type that trap was located in) intercept estimates for \\emph{Pterostichus melanarius}',col.names=c('Trap location','$\\beta$','S.E.','Z','p'),
      digits=c(1,2,2,2,2),align='rrrrl',escape=FALSE)
```

```{r PteMelSmooth, echo=FALSE, warning=FALSE}
pmelSmTerms %>% select(-Ref.df,-p) %>% 
  kable(.,caption='Smooth terms for \\emph{Pterostichus melanarius}. \\emph{s} indicates a thin-plate spline, \\emph{ti} indicates a tensor-product interaction. All terms except for day and (E,N) are functional regression fits.',
      col.names=c('Smoothing term','E.d.f.','$\\chi^2$','p'),
      digits=c(1,2,2,2),align='rrll',escape=FALSE)
```

```{r ParDisLinear, echo=FALSE, warning=FALSE}
pdisLinTerms %>% select(-p) %>% 
  kable(.,caption='Trap location (cover type that trap was located in) intercept estimates for \\emph{Pardosa distincta}',col.names=c('Trap location','$\\beta$','S.E.','Z','p'),
        digits=c(1,2,2,2,2),align='rrrrl',escape=FALSE)
```

```{r ParDisSmooth, echo=FALSE, warning=FALSE}
pdisSmTerms %>% select(-Ref.df,-p) %>% 
  kable(.,caption='Smooth terms for \\emph{Pardosa distincta}. \\emph{s} indicates a thin-plate spline, \\emph{ti} indicates a tensor-product interaction. All terms except for day and (E,N) are functional regression fits.',
        col.names=c('Smoothing term','E.d.f.','$\\chi^2$','p'),
        digits=c(1,2,2,2),align='rrll',escape=FALSE)
```

```{r ParMoeLinear, echo=FALSE, warning=FALSE}
pmoeLinTerms %>% select(-p) %>% 
  kable(.,caption='Trap location (cover type that trap was located in) intercept estimates for \\emph{Pardosa moesta}.',col.names=c('Trap location','$\\beta$','S.E.','Z','p'),
        digits=c(1,2,2,2,2),align='rrrrl',escape=FALSE)
```

```{r ParMoeSmooth, echo=FALSE, warning=FALSE}
pmoeSmTerms %>% select(-Ref.df,-p) %>% 
  kable(.,caption='Smooth terms for \\emph{Pardosa moesta}. \\emph{s} indicates a thin-plate spline, \\emph{ti} indicates a tensor-product interaction. All terms except for day and (E,N) are functional regression fits.',
        col.names=c('Smoothing term','E.d.f.','$\\chi^2$','p'),
        digits=c(1,2,2,2),align='rrll',escape=FALSE)
```

```{r OpilioLinear, echo=FALSE, warning=FALSE}
opilLinTerms %>% select(-p) %>% 
  kable(.,caption='Trap location (cover type that trap was located in) intercept estimates for \\emph{Phalangium opilio}',col.names=c('Trap location','$\\beta$','S.E.','Z','p'),
        digits=c(1,2,2,2,2),align='rrrrl',escape=FALSE)
```

```{r OpilioSmooth, echo=FALSE, warning=FALSE}
opilSmTerms %>% select(-Ref.df,-p) %>% 
  kable(.,caption='Smooth terms for \\emph{Phalangium opilio}. \\emph{s} indicates a thin-plate spline, \\emph{ti} indicates a tensor-product interaction. All terms except for day and (E,N) are functional regression fits.',
        col.names=c('Smoothing term','E.d.f.','$\\chi^2$','p'),
        digits=c(1,2,2,2),align='rrll',escape=FALSE)
```

# References

>>>>

---
title: "Non-crop sources of beneficial arthropods vary within-season across a prairie agroecosystem"
author:
  - name: Samuel V. J. Robinson
    email: samuel.robinson@ucalgary.ca
    affiliation: University of Calgary
    footnote: 1
  - name: Diane Edwards
    email: diane@abienvironmental.ca
    affiliation: Ambrose University
  - name: Jess L. Vickruck
    email: jessica.vickruck@canada.ca
    affiliation: University of Calgary
  - name: Lincoln R. Best
    email: lincoln.best@oregonstate.edu
    affiliation: Oregon State University
  - name: Paul Galpern
    email: paul.galpern@ucalgary.ca
    affiliation: University of Calgary
address:
  - code: University of Calgary
    address: 2500 University Drive NW, Calgary, AB
  - code: Ambrose University
    address: 150 Ambrose Circle SW, Calgary, AB
  - code: Oregon State University
    address: 1500 SW Jefferson Way, Corvallis, OR
footnote:
  - code: 1
    text: "Corresponding Author"
date: "Spring 2020"
abstract: "`r paste(readLines('abstract.Rmd'), collapse = ' ')`"
keep_tex: true
keywords:
  - beetles;
  - spiders;
  - harvestmen;
  - movement;
  - ecosystem services;
  - natural enemy;
  - functional data analysis;
  - seminatural land
bibliography: refs.bib
geometry: "left=3cm,right=3cm,top=3cm,bottom=3cm"
output: 
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    toc: false
    fig_caption: true
csl: journal-of-applied-ecology.csl    
header-includes:
  \makeatletter \def\ps@pprintTitle{  \let\@oddhead\@empty  \let\@evenhead\@empty  \def\@oddfoot{\centerline{\thepage}} \let\@evenfoot\@oddfoot} \makeatother
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{setspace}
  \linenumbers
---

@@

@@

## Data collection
We used a set of `r sum(NtrapsPer$n)` pitfall traps installed in road margins (minimum of 5 m away from the road edge, `r with(NtrapsPer, sum(n[trapLoc=='ditch']))` traps) and in-field locations (`r with(NtrapsPer, sum(n[trapLoc!='ditch']))`) across southern Alberta, Canada in 2017 (Figure \@ref(fig:siteMap)).
The sites spanned a west-to-east gradient of four natural subregions, including foothills parkland, foothills fescue, mixedgrass, and dry mixedgrass [@ABRegions2006].
Traps were placed starting on `r format(as.Date(as.character(min(PteMelMod$tempTrap$startjulian)),format='%j'),format='%B %d')`, and collections ended on `r format(as.Date(as.character(max(PteMelMod$tempTrap$endjulian)),format='%j'),format='%B %d')`, with collection occurring continuously and traps being emptied every `r round(mean(PteMelMod$tempTrap$trapdays),1)` days on average (SD: `r round(sd(PteMelMod$tempTrap$trapdays),1)`). 
This resulted in `r nrow(PteMelMod$tempTrap)` unique collection events, taking place across a total of `r round(sum(PteMelMod$tempTrap$trapdays),1)` trapping days.
In-field traps were placed in canola crops (`r with(NtrapsPer, sum(n[trapLoc=='canola']))`), wetlands (`r with(NtrapsPer, sum(n[trapLoc=='wetland']))`), grassy field edges (`r with(NtrapsPer, sum(n[trapLoc=='pivot']))`), and remnant prairie grasslands (`r with(NtrapsPer, sum(n[trapLoc=='native']))`).
Traps in canola were installed at 25, 75, and 200 m along a transect heading away from the nearest non-crop feature (wetland, grassy field edges, or remnant prairie), while the trap at 0 m was installed in the non-crop feature itself.
We used 582 mL Solo® cups buried up to the rim and partially filled with propylene glycol, with 2 cm wire mesh mounted over the rim to prevent vertebrates from falling into the traps.
Specimens were identified to species using appropriate taxonomic literature [@lindroth1966; @dondale1990; @edgar1990; @vogel2004; @yigit2007].

@@

Functional regression [@ramsay2004; @yen2014] was used to incorporate landscape information at different distances, allowing assessment of both local and regional landscape composition [@galpern2020].
Scalar-on-function regression is a special type of linear regression model ($y=X\beta + e$), where the columns of the model matrix $X$ contain some continuous predictor of the scalar $y$, and the values of the coefficients ($\beta$) are modeled as a smooth function $f(x)$ of the predictors.
<!-- As an example of scalar-on-function regression, biomass in a plot (scalar) could be modeled as a smoothed function of monthly average temperatures at that plot (function), meaning that the effect of monthly temperatures would also be a function of time. -->
In our case, the proportion of landscape cover ($X$) within each annulus surrounding the trap is a function of distance away from the trap, meaning that the coefficients are a smoothed function of distance ($\beta_i = f(distance_i)$), and represent the additive effect of a given type of landscape cover at distances away from the trap (i.e. $X\beta$).
This allows for the possibility that the size of the landscape "grain" relevant to a given organism may change over the course of its life stages [@addicott1987; @lima1996; @gardiner2010].
It also allows for the possibility that certain land cover types may be locally beneficial, but detrimental at large scales, indicating that the cover type may not constitute a completely usable habitat.

@@

@@

# Authors' contributions

SVJR and PG conceived of the project.
JV, DE, and LB conducted field work and taxonomic work.
SVJR designed the statistical analysis and wrote the manuscript.
JV, DE, LB, and PG all gave critical feedback on the manuscript, and give their approval for the final version.

@@

Many thanks to the farmers who gave us access to their land for this study. This project would not have been possible without the help of Michael Gavin, Jennifer Retzlaff, Mugunzal Abarmayan, Riley Waytes, Natasha Morden, and Amy Krause, who helped with field work, and John Swann, who providing lab space for sorting and identification. Funding for this research was provided by Ducks Unlimited Canada's Institute for Wetland and Waterfowl Research, the Alberta Canola Producers Commission, Manitoba Canola Growers Association, SaskCanola, the Alberta Biodiversity Monitoring Institute, and the Alberta Conservation Association.

# References {-}

<div id="refs"></div>

```{r child = 'supplemental.Rmd'}
```
